### 2024-07-29 mixed-models.qmd
20240729 - overview of mixed models

We will discuss a modeling framework that goes by many terms, including mixed models, mixed-effects models, multilevel models, hierarchical linear models.

>>>>

We will discuss a modeling framework that goes by many terms, including mixed models, mixed-effects models, multilevel models, hierarchical linear models.
They are sometimes called multilevel models and hiearchical linear models, whose name emphasizes the hierarchical structure of the data because the data are nonindependent.
When observations (i.e., data points) are collected from multiple lower-level units (e.g., people) in an upper-level unit (e.g., married couple, family, classroom, school, neighborhood, team), the data from the lower-level units are considered "nested" within the upper-level unit.
The data from the lower-level unit are likely to be correlated, to some degree, because they come from the same upper-level unit.
For example, multiple players may come from the same team, and the players' performance on that team is likely interrelated because they share common experiences and influence one another.
Thus, data from multiple players on a given team are considered nested within that team.
Longitudinal data can also be considered nested data, in which time points are nested within the person (i.e., the same player provides an observation across multiple time points).
As we will discuss, it is important to account for levels of nesting when the observations are nonindependent.

These models are also sometimes called mixed models or mixed-effects models because the models can include a mix of fixed and random effects.
Fixed effects are effects that are constant across individuals (i.e., upper-level units).
Random effects are effects that vary across individuals (i.e., upper-level units).
For instance, consider a longitudinal study of fantasy performance as a function of age.
If we have longitudinal data for multiple players, the time points are nested within players.
Examining the association between age as a fixed effect in relation to fantasy performace would examine the assocication between a player's age and their fantasy performance while holding the association between age and fantasy performance constant across all players.
That is, it would assume that all players show the same trajectory such as increase for 4 years then decrease.
Examining the association between age as a random effect in relation to fantasy performace would examine the assocication between a player's age and their fantasy performance while allowing the association between age and fantasy performance to vary across players.
That is, it would allow the possibility that some players improve with age, whereas other players decline in performance with age.

When including random effects of a variable (e.g., age) in a mixed model, it is also important to include fixed effects of that variable in the model.
This is because random effects have a mean of zero.
Fixed effects allow the mean to differ from zero.
Thus, inclusion of random effects without the corresponding fixed effect can lead to bias in estimation of the association between the predictor variables and the outcome variable.
